Staged Abstract Interpreters

### 2019-09-14 paper/50_stageabs.tex
polish

Now we can implement the staged abstract semantics, under that the binding-time
type @R@ is instantiated as @Rep@. The types of abstract components are reused
from the unstaged version.

@@

\paragraph{Primitive Operations} 
When deriving the staged concrete interpreter from its unstaged counterpart, we
notice that where the evaluation @ev(e)@ is invoked are shifted from @ap_clo@ to
@close@, where @e@ is the body expression of a $\lambda$-term.
In other words, when staging, we eagerly specialize the interpreter and
generate code every time we reach a $\lambda$-term, instead of lazily
calling @ev@ when the applications happen.  Similarly, here we use the same way
to handle staged $\lambda$-terms.  Additionally, in the staged abstract
interpreter, we have to handle nondeterminism incurred by the
over-approximation of runtime behavior.  In the rest of this part, we keep our
roadmap and focus on discussing the closure representation and function
application for the staged version.

The @close@ method now is a mix of the \textit{staged} concrete version and
unstaged \textit{abstract} version. We first build a current-stage function
@f@, which takes four next-stage values, including the argument and latest
store, and @in/out@ caches additionally. Inside @f@, we collapse the monadic
value of type @Ans@ to grounded values of type @Result@, by providing the
desired arguments, i.e., the new environment, new store, and two caches. The
collapsing of monadic values happens at the current-stage, so the invocation of
@ev@ is unfolded during staging. Finally, we generate a singleton set
containing the compiled next-stage closure (@emit_compiled_clo@), which is
represented by an IR node in LMS.
\begin{lstlisting}
  def emit_compiled_clo(f: (Rep[Value], Rep[Store], Rep[Cache], Rep[Cache])
                           => Rep[(Set[(Value,Store)], Cache)], λ: Lam, ρ: Exp[Env]): Rep[AbsValue]
  def close(ev: Expr => Ans)(λ: Lam, ρ: Rep[Env]): Rep[Value] = {
    val Lam(x, e) = λ
    val f: (Rep[Value],Rep[Store],Rep[Cache],Rep[Cache]) => Rep[(Set[(Value,Store)],Cache)] = {
      case (arg, σ, in, out) =>
        val α = alloc(σ, x)
        ev(e)(ρ + (unit(x) → α))(σ ⊔ Map(α → arg))(in)(out)
    }; Set[AbsValue](emit_compiled_clo(f, λ, ρ)) }
\end{lstlisting}

@@

\paragraph{Staged Caching and Fixpoint Iteration} The fixed-point iteration
again relies on the two caches @in@ and @out@, which are both staged maps now.
Therefore, the query of whether the @in/out@ cache contains the current
configuration will produce next-stage Boolean values, i.e., of type
@Rep[Boolean]@, and the branching condition cannot be determined statically.
We have to generate code for the whole @if@ expression.
Figure \ref{fig:staged_coind_cache} shows the staged version of @fix@. The
variable @res@ represents the next-stage result, consisting of a next-stage
@if@ expression. The true branch simply returns a pair of the query result
form the @out@ cache, and the @out@ cache itself. The else branch constructs a
monadic value @m@ of type @Ans@ first, which evaluates @e@ under the new @out@
cache.  After, we use a similar technique that collapses the monadic
value @m@ to grounded values, by providing its desired environment and etc.
Finally, we have a current-stage representation of the future-stage values
@res@, and we reify the content of @res@ back into the current-stage monad stack.

@@

\subsection{A Little Bit of Code Generation, Again}
The code generation for compiled closures and function applications are similar
to their counterparts in the staged concrete interpreter. We have two IR nodes
implemented as case classes; they also take additional caches as arguments. We
elide the code generation part for these IR nodes.
\begin{lstlisting}
case class IRCompiledClo(f: (Rep[Value], Rep[Store], Rep[Cache], Rep[Cache])
                          => Rep[(Set[ValSt], Cache)], λ: Lam, ρ: Rep[Env]) extends Def[AbsValue]
case class IRApClo(clo: Rep[AbsValue], arg: Rep[Value], σ: Rep[Store],
                    in: Rep[Cache], out: Rep[Cache]) extends Def[(Set[(Value, Store)], Cache)]
\end{lstlisting}

@@

Our staging schema works by unfolding the interpreter over the abstract syntax
tree of the input program. In practice, however, it would suffer from code
explosion when analyzing (specializing) large programs, which increases
compile time. If we generate next-stage programs running on the JVM, such large
generated programs would also incur GC overhead at runtime. In this
section, we present optimizations that largely mitigate these issues. 
Implementing all of those optimizations do not need to change the
generic interpreter.

@@

In the staged interpreters, all instances of @Env@ and @Store@ are staged. The
data structures representing these components are treated as black-boxes, i.e.,
@Rep[Map]@, which means that the operations on a @Rep[Map]@ directly become
next-stage code, and we do not inspect any further inside. As we identified when
introducing the generic interface, the keys of an @Env@ are string-represented
identifiers in the program, which are completely known statically. This
observation gives us a chance to further specialize the data structures for environments.
For example, let us assume that the @Map[K, V]@ is implemented as a hash-map.
If all the keys of type @K@ are all known statically, then the indices for
those keys can also be computed statically. Thus, the specialized map would be
an array of type @Array[Rep[V]]@, whose elements are next-stage values, and the
size of the array is known statically as the program has finite number of
identifiers. As result, an access to the map is staged into an access to the array
with the pre-computed index.

Particularly, if we are specializing a monovariant analysis, the address space
is equivalent to the set of all identifiers in the program. Utilizing this fact, the
result of accesses to the environment can be computed statically and we may
generate addresses directly during staging. After which, the store can be
specialized as an array by using the way mentioned above.

\paragraph{Selective Caching} We observe that the two-fold co-inductive
caching is used for every recursive call in our abstract interpreter. But this
is unnecessary and redundant when generating code for atomic
expressions such as literals or variables, because they always terminate.
Borrowing the idea from the partition of expressions used in A-Normal Form
$\lambda$-calculus \cite{Flanagan:1993:ECC:155090.155113}, we can use a
selective caching algorithm that does not generate caching code for atomic
expressions:
\begin{lstlisting}
  def fix_select: Expr => Ans = e => e match {
    case Lit(_) | Var(_) | Lam(_, _) => eval(fix_select)(e)
    case _ => fix_cache(e)
  }
\end{lstlisting}

\paragraph{Partially-static Data}
Our treatment of binding-times is coarse-grained: @Expr@s is static, the rest
of the components are all dynamic. But this is not always true, because the
static data have to be used somewhere with the dynamic operations.
Partial-static data is a way to improve binding-times and optimize the
generated code.
For example, to fold a static singleton set (often appears in @SetT@), e.g.,
@Set(x).foldLeft(init)(f)@ where @x@ and @init@ are staged values, a naive code
generator would faithfully generate code that applies @foldLeft@ with the set
and function @f@. But we can also utilize algebraic properties of @foldLeft@
to generate cheaper code, e.g., @Set(x).foldLeft(init)(f) = f(init, x)@. Since the
function @f@ is known at the current stage, we completely eliminate the fold
operation and function application. We apply several rewritings enabled by
partial-static patterns, such as for @Set@, @Map@, and @Tuple@. This
optimization greatly reduces the size of residual programs.

@@

We have gradually presented the confluence of specialization and abstraction of
concrete interpreters from an operational perspective.  In this section, we
review and summarize our recipe to achieve the staged abstract interpreter and
discuss correctness issues and different design choices.

@@

\begin{itemize}
  \item First, we construct a generic interpreter that abstracts over binding times,
    value domains, and primitive operations. In this paper, the generic interpreter
    is implemented in monadic style; therefore, the semantics can be encapsulated
    into monads.
  \item Then, we implement an unstaged abstract interpreter modularly using the
    appropriate monad stack. This step has been explored in the previous
    literatures.
  \item Finally, we replace the monad stack with a staged monad stack, and
    refactor related primitive operations.  Such staged monads operate on
    staged data types, i.e., next-stage values.
\end{itemize}

@@

Soundness is the central concern of static analysis, and as such, is vital for
prospective users. Our approach does not touch the soundness of the analysis,
i.e, if the unstaged one does not produces false negative result, the staged
one also does not. This soundness preservation indeed relies on the correctness
of the staged implementations and of the underlying MSP system. We now briefly
examine how this is achieved. Note that the rationale listed
here is based on empirical evidence; a rigorous proof for soundness
preservation of our approach remains an open challenge.

\begin{itemize}
  \item We usually assume the underlying MSP system preserves the meaning
    during staging, i.e., the generated program and original program produce
    the same result given the same arguments.
    In this paper, we use the LMS framework, which 1) checks binding-times by
    checking types which is enabled by the Scala compiler, and 2) generates
    code in A-Normal form \cite{Flanagan:1993:ECC:155090.155113} that preserves
    the execution order within a stage \cite{DBLP:conf/birthday/Rompf16}.
    However, in general, it is possible to subvert these guarantees by using
    unsafe features, such as casts.
  \item As shown in the previous sections, the generic interpreter is untouched
    and shared by the four artifacts.  This allows the programmer to
    check the correctness of the implementation of staged monads modularly.
   % We only instantiate binding-time annotations and the monadic types. 
  \item We build the staged monads and staged data structures in a
    correct-by-construct way that directly corresponds to their unstaged
    versions. For instance, the staged data structures we use here are simply
    black boxes that wrap the data structures in Scala library.
  \item In our experiments, the staged abstract interpreter produces the same
    result with the unstaged one on all benchmark programs we tested.
\end{itemize}

@@

\paragraph{Monadic-style vs Direct-style} We use a monadic interpreter
throughout the paper, but it is not necessary for staging. One can inline the
monadic operations and obtain an abstract interpreter in continuation-passing
style, or even translate back to a direct-style that may use explicit
side-effects such as mutations. In either case, we can still apply the staging
idea to the abstract interpreter and remove the interpretation overhead.
However, monads allow the staged abstract interpreter to be implemented in a
modular and extensible way.

>>>>

Now we can implement the staged abstract semantics, with the binding-time
type @R@ instantiated as @Rep@. The types of abstract components are reused
from the unstaged version.

@@

\paragraph{Primitive Operations} 
When deriving the staged concrete interpreter from its unstaged counterpart, we
notice that where the evaluation @ev(e)@ is invoked are shifted from @ap_clo@ to
@close@, where @e@ is the body expression of a $\lambda$-term.
In other words, when staging, we eagerly specialize the interpreter and
generate code every time we reach a $\lambda$-term, instead of lazily
calling @ev@ when the applications happen.  Similarly, here we use the same way
to handle staged $\lambda$-terms.  Additionally, in the staged abstract
interpreter, we have to handle nondeterminism incurred by the
over-approximation of runtime behavior.  In the rest of this part, we 
focus on discussing the closure representation and function
application for the staged version.

The @close@ method now is a mix of the \textit{staged} concrete version and
unstaged \textit{abstract} version. We first build a current-stage function
@f@, which takes four next-stage values, including the argument and latest
store, and @in/out@ caches additionally. Inside @f@, we collapse the monadic
value of type @Ans@ to grounded values of type @Result@, by providing the
desired arguments, i.e., the new environment, new store, and two caches. The
collapsing of monadic values happens at the current-stage, so the invocation of
@ev@ is unfolded at staging time. Finally, we generate a singleton set
containing the compiled next-stage closure (@emit_compiled_clo@), which is
represented by an IR node in LMS.
\begin{lstlisting}
  def emit_compiled_clo(f: (Rep[Value], Rep[Store], Rep[Cache], Rep[Cache])
                           => Rep[(Set[(Value,Store)], Cache)], λ: Lam, ρ: Exp[Env]): Rep[AbsValue]
  def close(ev: Expr => Ans)(λ: Lam, ρ: Rep[Env]): Rep[Value] = {
    val Lam(x, e) = λ
    val f: (Rep[Value],Rep[Store],Rep[Cache],Rep[Cache]) => Rep[(Set[(Value,Store)],Cache)] = {
      case (arg, σ, in, out) =>
        val α = alloc(σ, x)
        ev(e)(ρ + (unit(x) → α))(σ ⊔ Map(α → arg))(in)(out)
    }; Set[AbsValue](emit_compiled_clo(f, λ, ρ)) }
\end{lstlisting}

@@

\paragraph{Staged Caching and Fixpoint Iteration} The fixed-point iteration
again relies on the two caches @in@ and @out@, which are both staged maps now.
Therefore, the query of whether the @in/out@ cache contains the current
configuration will produce next-stage Boolean values, i.e., of type
@Rep[Boolean]@, and the branching condition cannot be determined statically.
We have to generate code for the whole @if@ expression.
Figure \ref{fig:staged_coind_cache} shows the staged version of @fix@. The
variable @res@ represents the next-stage result, consisting of a next-stage
@if@ expression. The true branch simply returns a pair of the query result
form the @out@ cache, and the @out@ cache itself. The else branch constructs a
monadic value @m@ of type @Ans@ first, which evaluates @e@ under the new @out@
cache.  After this, we use a similar technique that eagerly collapses the
monadic value @m@ to grounded values, by providing its desired environment and
other arguments.  Finally, we have a current-stage representation of the
future-stage values @res@, and we reify the content of @res@ back into the
current-stage monad stack.

@@

\subsection{A Little Bit of Code Generation, Again}
The code generation for compiled closures and function applications is similar
to their counterparts in the staged concrete interpreter. We have two IR nodes
implemented as case classes; they also take additional caches as arguments. We
elide the code generation part for these IR nodes.
\begin{lstlisting}
case class IRCompiledClo(f: (Rep[Value], Rep[Store], Rep[Cache], Rep[Cache])
                          => Rep[(Set[ValSt], Cache)], λ: Lam, ρ: Rep[Env]) extends Def[AbsValue]
case class IRApClo(clo: Rep[AbsValue], arg: Rep[Value], σ: Rep[Store],
                    in: Rep[Cache], out: Rep[Cache]) extends Def[(Set[(Value, Store)], Cache)]
\end{lstlisting}

@@

Our staging schema works by unfolding the interpreter over the abstract syntax
tree of the input program. In practice, however, the staging schema would
suffer from code explosion when analyzing (specializing) large programs, which
increases compile time. If we generate next-stage programs running on the JVM,
such large generated programs would also incur GC overhead at runtime. In this
section, we present optimizations that largely mitigate these issues.
Implementing all of those optimizations do not need to change the generic
interpreter.

@@

In the staged interpreters, all instances of @Env@ and @Store@ are staged. The
data structures representing these components are treated as black-boxes, i.e.,
@Rep[Map]@, which means that the operations on a @Rep[Map]@ directly become
next-stage code, and we do not inspect any further inside. As we identified when
introducing the generic interface, the keys of an @Env@ are string-represented
identifiers in the program, which are completely known statically. This
observation gives us a chance to further specialize the data structures for environments.
For example, let us assume that the @Map[K, V]@ is implemented as a hash-map.
If all the keys of type @K@ are all known statically, then the indices for
those keys can also be computed statically. Thus, in this case, the specialized
map will be an array of type @Array[Rep[V]]@, whose elements are next-stage
values, and the size of the array is known statically as the program has finite
number of identifiers. As result, an access to the map is staged into an access
to the array with a pre-computed index.

In particular, if we are specializing a monovariant analysis, the address space
is equivalent to the set of all identifiers in the program. Utilizing this fact, the
result of accesses to the environment can be computed statically and we may
generate addresses directly during staging. After this, the store can be
specialized as an array by using the way mentioned above.

\paragraph{Selective Caching} We observe that the two-fold co-inductive
caching is used for every recursive call in our abstract interpreter. But this
is unnecessary and redundant when generating code for atomic
expressions such as literals or variables, because they always terminate.
Borrowing the idea from the partition of expressions used in 
administrative normal form (ANF) $\lambda$-calculus
\cite{Flanagan:1993:ECC:155090.155113}, we can use a selective caching
algorithm that does not generate caching code for atomic expressions:
\begin{lstlisting}
  def fix_select: Expr => Ans = e => e match {
    case Lit(_) | Var(_) | Lam(_, _) => eval(fix_select)(e)
    case _ => fix_cache(e)
  }
\end{lstlisting}

\paragraph{Partially-static Data}
Our treatment of binding-times is coarse-grained: @Expr@s are static, the rest
of the components are all dynamic. But this is not always true, because the
static data have to be used somewhere with the dynamic operations.
Partially-static data is a way to improve binding-times and optimize the
generated code.
For example, to left-fold a static singleton set (often appears in @SetT@), e.g.,
@Set(x).fold(init)(f)@ where @x@ and @init@ are staged values, a naive code
generator would faithfully generate code that applies @fold@ with the set
and function @f@. But we can also utilize algebraic properties of @fold@
to generate cheaper code, e.g., @Set(x).fold(init)(f) = f(init, x)@. Since the
function @f@ is known at the current stage, we completely eliminate the fold
operation and function application. We apply several rewritings enabled by
partially-static patterns, such as for @Set@, @Map@, and @Tuple@. This
optimization greatly reduces the size of residual programs.

@@

We have gradually presented the confluence of specialization and abstraction of
concrete interpreters from an operational perspective.  In this section, we
review and summarize our recipe to achieve the staged abstract interpreter and
discuss correctness and different design choices.

@@

\begin{itemize}
  \item First, we construct a generic interpreter that abstracts over binding times,
    value domains, and primitive operations. In this paper, the generic interpreter
    is implemented in monadic style; therefore, the semantics can be encapsulated
    into monads.
  \item Then, we implement an unstaged abstract interpreter modularly using the
    appropriate monad stack. This step has been explored in the previous
    literature.
  \item Finally, we replace the monad stack with a staged monad stack, and
    refactor related primitive operations.  Such staged monads operate on
    staged data types, i.e., next-stage values.
\end{itemize}

@@

Soundness is the central concern of static analysis, and as such, is vital for
prospective users. Our approach does not interact with the soundness of the analysis,
i.e, if the unstaged one does not produces false negative result, the staged
one also does not. This soundness preservation indeed relies on the correctness
of the staged implementations and the underlying MSP system. We now briefly
examine how this is achieved. Note that the rationale listed
here is based on empirical evidence; a rigorous proof of soundness
preservation of our approach remains an open challenge.

\begin{itemize}
  \item 
    In general, syntactic MSP systems based on quotations (e.g.,
    MetaML/MetaOCaml) cannot guarantee that the generated code preserves the
    meaning after staging. In other words, the generated code may not preserve
    the evaluation order or may contain undesired duplication of code.
    In this paper, we use the LMS framework, which 1) checks binding-times by
    checking types which is enabled by the Scala compiler, and 2) generates
    code in administrative normal form \cite{Flanagan:1993:ECC:155090.155113}
    that preserves the execution order within a stage
    \cite{DBLP:conf/birthday/Rompf16}.
    %However, in general, it is possible to subvert these guarantees by using
    %unsafe features, such as casts.
  \item As shown in the previous sections, the generic interpreter is untouched
    and shared by the four artifacts.  This allows the programmer to
    check the implementation of staged monads and primitive operations,
    as well as their correctness modularly.
   % We only instantiate binding-time annotations and the monadic types. 
  \item We build the staged monads and staged data structures in a
    correct-by-construct way that directly corresponds to their unstaged
    versions. For instance, the staged data structures we use here are simply
    black boxes that wrap the data structures in the Scala library.
  \item In our experiments, the staged abstract interpreter produces the same
    result as the unstaged one on all benchmark programs we tested.
\end{itemize}

@@

\paragraph{Monadic-style vs Direct-style} We use a monadic interpreter
throughout the paper, but using monadic-style is not the only choice for
staging. One can inline the monadic operations and obtain an abstract
interpreter in continuation-passing style, or even translate back to a
direct-style that may use explicit side-effects such as mutations. In either
case, we can still apply the staging idea to the abstract interpreter and
remove the interpretation overhead.  However, monads allow the staged abstract
interpreter to be implemented in a modular and extensible way.

### 2019-09-14 paper/60_cases.tex
polish

In Section~\ref{sai}, we use a toy language to show that staging an abstract
interpreter is feasible and the artifact can be derived by combining the staged
concrete interpreter and unstaged abstract interpreter.
In this section, we conduct several case studies to further show that this
methodology is practically useful and widely applicable to diverse analyses.

@@

Specifically, \citeauthor{Boucher:1996:ACN:647473.727587} showed how to compile a
monovariant control-flow analysis \cite{Shivers:1991:SSC:115865.115884,
Shivers:1988:CFA:53990.54007} for continuation-passing style (CPS) programs.
The analyzer is a big-step control-environment abstract interpreter. 
By applying AC, we refactor every function in the analyzer such that it
returns a closure taking an environment argument.  Therefore the overhead
of traversing the AST of input program is eliminated.

We show that AC can be understood and implemented as an instance of staging abstract
interpreters. We first revisit the original implementation of abstract
compilation for 0-CFA, and then reproduce their result by simply adding stage
annotations.  The program generated by our approach provides approximately the
same improvement with regards to speed, but without changing a single line of
the analyzer program using type-based staging annotations. However, AC requires
more engineering effort, i.e., rewriting the whole analyzer into the
closure-generation form. Moreover, as shown in Section~\ref{staged_ds}, our
approach is able to not only remove the interpretive overhead, but also enable
several optimizations such as specialized data structures.

@@

To demonstrate that the SAI approach is applicable to various analyses, we now
extend and refactor the staged abstract interpreter we presented in Section
\ref{sai} to several variants of control-flow analyses (CFA). We first discuss
the calling-context sensitive analysis for staged analyzers by tweaking the
allocation strategy. Then, we describe how to implement store-widening with the
staged monads, which is a common optimization in CFA. Finally, we integrate
abstract garbage collection with staging. All of them are orthogonal to
staging.

@@

The abstract addresses we introduced in Section \ref{sai} are simply identical
to the variable names, which is a context-insensitive schema.  Through tweaking
the address allocation strategy, we can achieve various context-sensitive
analyses \cite{DBLP:conf/icfp/Gilray0M16}. Here, we demonstrate a
calling-context sensitive analysis, i.e., $k$-CFA-like analysis
\cite{DBLP:journals/jfp/HornM12}, can be implemented in our staged abstract
interpreter.
\begin{lstlisting}
  type Time = List[Expr]
  case class KCFAAddr(x: Ident, time: Time) extends Addr
\end{lstlisting}

The timestamp is essentially a finite list of expressions that tracks the $k$-most
recent calling contexts. The definition of abstract addresses @KCFAAddr@ is also
changed to include an identifier and the time when it gets allocated, meaning that
this address points to some values under such calling context. If $k$ is 0, we
obtain a monovariant analysis as demonstrated before; if $k > 0$, we obtain a
family of analyses with increasing precision. Note that when $k = 0$, the
address space is statically determined by the set of variables appeared in the
program, and we may exploit it to optimize the generated code. But when $k > 0$,
the address generation happens at the next stage, i.e., analysis time.
\begin{lstlisting}
  type AnsM[T] = RepReaderT[RepStateT[RepStateT[ // add Time with a RepStateT
                  RepSetReaderStateM[Cache, Cache, ?], Time, ?], Store, ?], Env, T]
  type Result = (Rep[Set[(Value, Store, Time)]], Rep[Cache])
\end{lstlisting}

We then integrate the timestamp into our monad stack by adding a staged
@StateT@ monad. The result type is accordingly updated to contain the set of
tuples, where the time is added.
\begin{lstlisting}
  def tick(e: Expr): AnsM[Unit] = for {
    τ <- get_time; u <- liftM(StateTMonad.put((e::τ).take($k$)))
  } yield u
  def alloc(x: String): AnsM[Addr] = for {
    τ <- get_time
  } yield unit(KCFAAddr(x, τ)) // we use unit to turn a current-stage constant to a next-stage value
\end{lstlisting}

@@

A commonly-used technique to improve the running time of control-flow analyses
is store-widening. As shown by \cite{Darais:2015:GTM:2814270.2814308,
DBLP:journals/pacmpl/DaraisLNH17}, a store-widening analysis can be easily
implemented in the monadic abstract interpreter by swapping the @StateT@ and
the @SetT@ monad transformer. This optimization is orthogonal to staging.
Within the monadic interpreter framework, we just need to similarly swap the
same \textit{staged} version of monads. After this adjustment, conceptually, we
have the following monad stack:
\begin{lstlisting}
  type AnsM[T] = RepReaderT[RepSetT[RepStateT[
                   RepReaderT[RepStateT[IdM, Cache, ?], Cache, ?], Store, ?], ?], Env, T]
\end{lstlisting}

@@

Abstract garbage collection is a technique to reclaim unreachable addresses in
the store while doing the abstract interpretation
\cite{Might:2006:IFA:1159803.1159807}. \citet{DBLP:journals/pacmpl/DaraisLNH17}
showed that, for big-step monadic abstract interpreters, we may add a @Reader@
monad to track the set of root addresses and compute the reachable addresses every
time when we obtain a value. Here, we adopt the idea with staged monads:
\begin{lstlisting}
  type AnsM[T] = RepReaderT[RepReaderT[
                   RepStateT[RepSetReaderStateM[Cache, Cache, ?], Store, ?],
                 Set[Addr], ?], Env, T] //add a reader monad for Set[Addr]
\end{lstlisting}

The skeleton of a staged version of abstract garbage collection implementation
is shown in Figure \ref{fixgc}.
We refactor the @fix@ function with caching (used in Section \ref{sai}) to @fix_gc@
(the caching part is omitted in the presented code snippet). Several auxiliary
functions are used: @ask_root@ retrieves the current root addresses from the
monad stack; @root@ computes the reachable addresses given a value; and @gc@
performs garbage collection according to the given reachable addresses and
returns a new store. We also need to compute the root addresses for compound
expressions, which is elided in the code snippet; for the details of this part,
readers may refer to \cite{DBLP:journals/pacmpl/DaraisLNH17}.

@@

One of the challenges of modern static analysis is that programs usually depend
on large libraries, and analyzing these libraries is time-consuming
\cite{toman_et_al:LIPIcs:2017:7121}.
If we are able to analyze programs and libraries separately and reuse the
results without losing precision, we can reduce the overhead of repeatedly
analyzing libraries. Some static analyzers compute a summary for a function or
a module, which can be reused later. However, when lacking the context
information, those analyses can be too conservative or unsound, which both lead
to imprecision. In this section, we show that meta-programming can provide a
trade-off between summary-based modular analyses and slow whole-program
analyses, by utilizing an existing whole-program analyzer.

As we have already seen in Section \ref{sai}, the primitive operation @close@
can denote a $\lambda$-term to a next-stage Scala function. Similar to
compilers, the specializing of the staged abstract interpreter to a
$\lambda$-term is also independent of other parts of the analyzed program.
\begin{lstlisting}
  def close(ev: Eval => Ans)(λ: Expr, ρ: Rep[Env]): Rep[Value]
\end{lstlisting}

>>>>

In Section~\ref{sai}, we use a toy language to show that staging an abstract
interpreter is feasible and that a staged artifact can be derived by combining
the staged concrete interpreter and unstaged abstract interpreter.
In this section, we conduct several case studies to further show that this
methodology is practically useful and widely applicable to a diverse set of
analyses.

@@

Specifically, \citeauthor{Boucher:1996:ACN:647473.727587} showed how to compile a
monovariant control-flow analysis \cite{Shivers:1991:SSC:115865.115884,
Shivers:1988:CFA:53990.54007} for continuation-passing style (CPS) programs.
The analyzer is a big-step control-environment abstract interpreter. 
By applying AC, every function in the analyzer is refactored to return a
closure only taking an environment argument.  Therefore the overhead of
traversing the AST of input program is eliminated.

We show that AC can be understood and implemented as an instance of staging abstract
interpreters. We first revisit the original implementation of abstract
compilation for 0-CFA, and then reproduce their result by simply adding stage
annotations.  The program generated by our approach provides approximately the
same improvement with regards to speed, but without changing a single line of
the analyzer program relying on type-based staging annotations. With Scala's
type inference, we only need to add staging annotations for function argument
types and return types.
However, AC requires more engineering effort, i.e., rewriting the whole
analyzer into the closure-generation form. Moreover, as shown in
Section~\ref{staged_ds}, our approach is able to not only remove the
interpretive overhead, but also enable several optimizations such as
specialized data structures.

@@

To demonstrate that the SAI approach is applicable to various analyses, we now
extend and refactor the staged abstract interpreter we presented in Section
\ref{sai} to several variants of control-flow analyses (CFA). We first discuss
the calling-context sensitive analysis for staged analyzers by tweaking the
allocation strategy. Then, we describe how to implement store-widening with the
staged monads, which is a common optimization in CFA. Finally, we integrate
abstract garbage collection with staging. All of these extensions are orthogonal to
staging.

@@

The abstract addresses we introduced in Section \ref{sai} are identical
to the variable names, which is a context-insensitive schema.  Through tweaking
the address allocation strategy, we can achieve various context-sensitive
analyses \cite{DBLP:conf/icfp/Gilray0M16}. Here, we demonstrate a
calling-context sensitive analysis, i.e., $k$-CFA-like analysis
\cite{DBLP:journals/jfp/HornM12}, can be implemented in our staged abstract
interpreter.
\begin{lstlisting}
  type Time = List[Expr]
  case class KCFAAddr(x: Ident, time: Time) extends Addr
\end{lstlisting}

The timestamp is a finite list of expressions that tracks the $k$-most
recent calling contexts. The definition of abstract addresses @KCFAAddr@ is also
changed to include an identifier and the time when it gets allocated, meaning that
this address points to some values under such calling context. If $k$ is 0, we
obtain a monovariant analysis as demonstrated before; if $k > 0$, we obtain a
family of analyses with increasing precision. Note that when $k = 0$, the
address space is statically determined by the set of variables appeared in the
program, and we may exploit it to optimize the generated code. But when $k > 0$,
the address generation happens at the next stage, i.e., analysis time.
\begin{lstlisting}
  type AnsM[T] = RepReaderT[RepStateT[RepStateT[ // add Time with a RepStateT
                  RepSetReaderStateM[Cache, Cache, ?], Time, ?], Store, ?], Env, T]
  type Result = (Rep[Set[(Value, Store, Time)]], Rep[Cache])
\end{lstlisting}

We then integrate the timestamp into our monad stack by adding a staged
@StateT@ monad. The result type is accordingly updated to contain the set of
tuples, where the field of @Time@ is added.
\begin{lstlisting}
  def tick(e: Expr): AnsM[Unit] = for {
    τ <- get_time; u <- liftM(StateTMonad.put((e::τ).take($k$)))
  } yield u
  def alloc(x: String): AnsM[Addr] = for {
    τ <- get_time
  } yield unit(KCFAAddr(x, τ)) // we use unit to turn a current-stage constant to a next-stage value
\end{lstlisting}

@@

A commonly-used technique to improve the running time of control-flow analyses
is store-widening. As shown by \citet{Darais:2015:GTM:2814270.2814308,
DBLP:journals/pacmpl/DaraisLNH17}, a store-widening analysis can be easily
implemented in the monadic abstract interpreter by swapping the @StateT@ and
the @SetT@ monad transformers. This optimization is orthogonal to staging.
Within the monadic interpreter framework, we just need to swap the
same \textit{staged} version of monad transformers. After this adjustment, conceptually, we
have the following monad stack:
\begin{lstlisting}
  type AnsM[T] = RepReaderT[RepSetT[RepStateT[
                   RepReaderT[RepStateT[IdM, Cache, ?], Cache, ?], Store, ?], ?], Env, T]
\end{lstlisting}

@@

Abstract garbage collection is a technique to reclaim unreachable addresses in
the store while performing the abstract interpretation
\cite{Might:2006:IFA:1159803.1159807}. \citet{DBLP:journals/pacmpl/DaraisLNH17}
showed that, for big-step monadic abstract interpreters, we may add a @Reader@
monad to track the set of root addresses and compute the reachable addresses every
time when we obtain a value. Here, we adopt the idea with staged monads:
\begin{lstlisting}
  type AnsM[T] = RepReaderT[RepReaderT[
                   RepStateT[RepSetReaderStateM[Cache, Cache, ?], Store, ?],
                 Set[Addr], ?], Env, T] //add a reader monad for Set[Addr]
\end{lstlisting}

The skeleton of a staged version of abstract garbage collection implementation
is shown in Figure \ref{fixgc}.
We refactor the @fix@ function with caching (used in Section \ref{sai}) to @fix_gc@
(the caching part is omitted in the presented code snippet). Several auxiliary
functions are used: @ask_root@ retrieves the current root addresses from the
monad stack; @root@ computes the reachable addresses given a value; and @gc@
performs garbage collection according to the given reachable addresses and
returns a new store. We also need to compute the root addresses for compound
expressions, which is elided in the code snippet; for the details of this part,
readers may refer to \citet{DBLP:journals/pacmpl/DaraisLNH17}.

@@

One of the challenges of modern static analysis is that programs usually depend
on large libraries, and analyzing these libraries is time-consuming
\cite{toman_et_al:LIPIcs:2017:7121}.
If we are able to analyze programs and libraries separately and reuse the
results without losing precision, we can reduce the overhead of repeatedly
analyzing libraries. Some static analyzers compute a summary for a function or
a module, which can be reused later. However, when lacking context
information, those analyses can be too conservative or unsound. 
In this section, we show that meta-programming can provide a
trade-off between summary-based modular analyses and slow whole-program
analyses, by utilizing an existing whole-program analyzer.

As we have already seen in Section \ref{sai}, the primitive operation @close@
can denote a $\lambda$-term to a next-stage Scala function. 
The specialization of the staged abstract interpreter to a
$\lambda$-term is also independent of other parts of the analyzed program.
\begin{lstlisting}
  def close(ev: Eval => Ans)(λ: Expr, ρ: Rep[Env]): Rep[Value]
\end{lstlisting}
